### 2021-05-31 paper/parlr_paper_1.tex
wss slides first draft

\hypertarget{introduction-or-abstract}{%
\subsection{Introduction (or
Abstract?)}\label{introduction-or-abstract}}

Record linkage is the task of identifying duplicate records across
multiple data sources. This task is straightforward when provided unique
identifiers or highly reliable information, but becomes difficult when
data sources are plagued with error. Additionally, record linkage is
inherently computationally intensive, and many existing methods are not
feasible for many applications. In this paper, we propose a Bayesian
model for record linkage that is amenable to parallel computing, hashing
techniques and efficient storage mechanisms, that is scalable to large
linkage tasks.

@@

Two modern adaptations of the Fellegi Sunter model are important for
understanding \texttt{parlr} and its contribution. In 2019, Enamorado et
al proposed \texttt{fastlink}, a method that closely followed the
modelling assumptions of Fellegi and Sunter, but used innovating hashing
techniques to greatly enhance its speed and scale. Importantly, they
maintained Fellegi and Sunter's independent matching assumption, in
which the matching status of any \((i,j)\) pair is made independently of
the matching of any other \((i', j')\). This leads to many matchings
that violate ``one-to-one'' considerations that are often implicit in
the data, which need to be resolved in a post-processing step. In
contrast, Sadinle (2017) proposed ``Beta Record Linkage for Bipartite
Matching'' and the accompanying \texttt{BRL} package which strictly
enforces one-to-one matching. Specifically, in each iteration of his
Gibbs sampler, he considers each record \(j\in B\), removes from
consideration the records \(i\in A\) that have already been matched,
then samples a potential link. However, accounting for these
dependencies throughout the linkage process is computationally
burdensome, leaving \texttt{BRL} only suitable for small to moderate
linkage problems.

Outside of the Fellegi-Sunter framework, some researchers have developed
methods that model the data directly, not a derived set of comparison
vectors. As an early example, Steorts et al 2016 produced
\texttt{blink}, which had the advantage of a computational complexity
that grew linearly with the data (as opposed to quadratically within
Fellegi Sunter), but was plagued with slow mixing times and was unable
to incorporate text data. Later, Steorts et al 2020 improved on this
method with \texttt{dblink}, which was able to simulate text data within
its Gibbs sampler by drawing from an empirical distribution, and
remedied its slow mixing through a probabilistic blocking approach.
However, this method remains computationally intensive and requires
distributed computing to manage even moderate linkage problems in
acceptable time.

@@

Our notation and assumptions closely follow that of Sadinle (2017).
Denote two files as \(A\) and \(B\), with \(n_A\) and \(n_B\) records
respectively, and with records indexed as \(i \in \{1, \ldots, n_A\}\)
in \(A\) and \(j \in \{1, \ldots, n_B\}\) in \(B\). Without loss of
generality, label the files such that \(n_A \geq n_B\). We also assume
there are no duplicates within files, only across. For each record pair
under consideration, we generate a comparison vector
\(\boldsymbol{\gamma}_{ij} = \{\gamma_{ij}^1, \ldots, \gamma_{ij}^F\}\),
where \(F\) is the number of fields used in the linkage and each
\(\gamma_{ij}^f\) takes on a value \(l \in \{1, \ldots, L_f\}\)
indicating the level agreement between the two records on a specified
field.

@@

\hypertarget{model-specification}{%
\section{Model Specification}\label{model-specification}}

@@

Here the reader should the relationship between our proposed model and
that of Sadinle 2017. In his model, Sadinle constructs a prior
distribution on the entire \(\mathbf{Z}\) vector, which induces a Gibbs
sampler that strictly enforces one-to-one matching. In particular, this
sampler removes previously matches records from the set of candidate
records when sampling \(Z_j\), creating a dependency that makes the
sampler \emph{inherently serial}. We however use independent priors for
each \(Z_j\), thereby weakening the one-to-one requirement; our sampler
does ensure that each record in \(B\) can be matched to at most one
record in \(A\), but allows for the possiblity that multiple records in
\(B\) match to the same record in \(A\). With this change, components of
the \(\mathbf{Z}\) vector can be computed in parallel, allowing for
significant computational gains. More importantly, since only the
agreement pattern of \(Z_j\) is used for calculations within the Gibbs
sampler, and not the particular record label, we can conduct this
sampling only at the level of the unique agreement patterns, offering
even more computational savings. This will be explained further in
\textcolor{red}{Section XX}.

@@

To address this, we propose a new strategy called \emph{storage
efficient indexing} (SEI). In standard indexing, the modeler decides a
certain criteria that they expect all true matching pairs to satisfy,
and a priori label any record pairs that do not meet that criteria as
non-matching. For example, one might only consider pairs with a certain
similarity score on a field deemed to be important (like first name), or
only pairs with exact matching on a specified number of fields. While
generally chosen to be be quite loose, establishing these criteria
requires knowledge of the problem and invites room for human error. We
propose a method of reducing the comparison space (and reducing storage
requirements) without these drawbacks. Note that all records of the same
agreement pattern have the same probability when sampling \(Z_j\).
Therefore we know that records belonging to an \(h_p\) such that
\(H_{p_j}\) is large are very unlikely to be sampled consistently enough
to be deemed a match through the Bayes estimate, even without
considering the form of the agreement pattern itself. In random
indexing, rather than store all of these unlikely record labels, we
choose to store only a small number \(R\) of them. Posterior
calculations still attribute the appropriate weight to all records
through the summary statistics \(H_{p_j}, H_p^m\), and \(H_p^u\); this
is similar to the \emph{u-correction} concept introduced by
\_\_\_\_\_\_\_\_. Rather than storing \(n_A \times n_B\) record labels,
random indexing allows us to store at most \(n_B \times P \times R\),
regardless of how large \(n_A\) is. Coupled with the partitioning
strategy for creating comparison vectors, this greatly expands the size
of the linkage tasks we can undertake.

@@

\hypertarget{simulation-studies}{%
\section{Simulation Studies}\label{simulation-studies}}

In his 2017 paper, Sadinle demonstrated the strength of his method by
running \texttt{BRL} on simulated pairs of files with differing levels
of errors and overlap. Comparing our method against \texttt{BRL} on
these same simulated datasets, we find that our method only has weakened
performance in the extreme scenario of very high errors and very high
overlap across files. We use the following measures of accuracy in our
analysis: although using only recall and precision is common in the
literature, we find it useful to have a single metric the gauge overall
performance, and thus also include F measure.

@@

To demonstrate speed, we generate comparison vectors from pre-specified
distributions so that we can easily increase the size of the linkage
problem. Distributions are meant to emulate the behavior of similarity
scores across first name, last name, and day, month, and and year of
birth, with 1 indicating an exact match on a particular field and 0
indicating a nonmatch. We simulate these data for different values of
\(n_A\) and \(n_B\), and compare the run-time of \texttt{parlr} against
\texttt{BRL}. We see that at low data size, \texttt{BRL} outperforms,
but that \texttt{parlr} is significantly faster at handling larger data.

In particular, run-time for \texttt{BRL} seems to grow quadratically
while run-time for \texttt{parlr} seems to grow linearly.

@@

The above discussion suggests that for fixed \(n_B\), computation time
should remain mostly constant with growing \(n_A\). Shockingly, this
seems to be true. In the plot below, fixing \(n_B = 500\), we see linear
growth for the run-time under \texttt{BRL} as \(n_A\) increases, with
much more static run-time under \texttt{parlr}. The slight increases in
run-time that we do see are due primarily to the hashing step, which
again can be run in parallel for large data.

@@

\begin{itemize}
\tightlist
\item
  Thinking of re-running the El Salvador linkage from the Sadinle 2017
  paper, but doing so with without the formal blocking and indexing that
  he uses. Instead, I'd use hashing and storage efficient indexing -- I
  should be able to do linkage on the all the data at once!
\end{itemize}

>>>>

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Record linkage is the task of identifying duplicate records across
multiple data sources. This task is straightforward when provided unique
identifiers or highly reliable information, but becomes difficult when
data sources are plagued with error. Additionally, many record linkage
tasks are large in nature, and many existing methods are too
computational intensive to be used in practice. In this paper, we
propose a Bayesian model for record linkage that is amenable to parallel
computing, hashing techniques and efficient storage mechanisms, that is
scalable to large linkage tasks.

The proposed method is inspired by the ``Beta Record Linkage'' method
proposed by Mauricio Sadinle's 2017 ``Bayesian Estimation of Bipartite
Matching for Record Linkage''. As such, we demonstrate our method
through a comparison of our methods against Sadinle's on the simulation
study he designed and the El Salvador homocide case study he utilized.
We include one additional simulation study to demonstrate the power of
our method on large datasets. In what follows, Section 2 describes
previous work in record linkage that lays the foundation for our method,
Section 3 provides our model and the Gibbs sampler used for posterior
inference, and Section 4 describes strategies used for more efficient
and scalable implementation. Then, Sections 5 and 6 demonstrate the
method through simulations and a case study. Lastly, our discussion
highlights open questions in the record linkage liturature and motivates
future work.

@@

The likelihood in the original Fellegi Sunter model is a mixture model
through which each record pair is independently classified as a match or
nonmatch. This independent classification often leads to sets of matches
that break expectations of transisitivity, and Jaro devised an procedure
to reduce a set of conflicting matches to the mostly likely set of
one-to-one matchings. One key strength of Sadinle's \texttt{BRL} is that
it is garunteed to produce a set of bipartite matchings that respect
one-to-one assumptions--without the need for any post-hoc procedures.
Specifically, in each iteration of his Gibbs sampler, he considers each
record \(j\in B\), removes from consideration the records \(i\in A\)
that have already been matched, and then samples a potential link.
However, accounting for these dependencies throughout the linkage
process is computationally burdensome, leaving \texttt{BRL} only
suitable for small to moderate linkage problems.

Outside of the Fellegi-Sunter framework, some researchers have developed
methods that model the data directly, not a derived set of comparison
vectors. As an early example, Steorts et al 2016 produced
\texttt{blink}, which had the advantage of a computational complexity
that grew linearly with the data (as opposed to quadratically within
Fellegi Sunter), but was plagued with slow mixing times and was unable
to incorporate text data. Later, Steorts et al 2020 improved on this
method with \texttt{dblink}, which was able to simulate text data within
its Gibbs sampler by drawing from an empirical distribution, and
remedied its slow mixing through a probabilistic blocking approach.
However, this method remains computationally intensive and requires
distributed computing to manage even moderate linkage problems in
acceptable time. Because these methods fall outside of the
Fellegi-Sunter framework, we do not consider them further in this paper.

@@

Our notation and assumptions closely follow those of Sadinle. Denote two
files as \(A\) and \(B\), with \(n_A\) and \(n_B\) records respectively,
and with records indexed as \(i \in \{1, \ldots, n_A\}\) in \(A\) and
\(j \in \{1, \ldots, n_B\}\) in \(B\). Without loss of generality, label
the files such that \(n_A \geq n_B\). We also assume there are no
duplicates within files, only across. For each record pair under
consideration, we generate a comparison vector
\(\boldsymbol{\gamma}_{ij} = \{\gamma_{ij}^1, \ldots, \gamma_{ij}^F\}\),
where \(F\) is the number of fields used in the linkage and each
\(\gamma_{ij}^f\) takes on a value \(l \in \{1, \ldots, L_f\}\)
indicating the level agreement between the two records on a specified
field.

@@

\hypertarget{model-specification}{%
\subsection{Model Specification}\label{model-specification}}

@@

Here the reader should the relationship between our proposed model and
that of Sadinle 2017. In his model, Sadinle constructs a prior
distribution on the entire \(\mathbf{Z}\) vector, which induces a Gibbs
sampler that strictly enforces one-to-one matching. In particular, this
sampler removes previously matches records from the set of candidate
records when sampling \(Z_j\), creating a dependency that makes the
sampler \emph{inherently serial}. We however use independent priors for
each \(Z_j\), thereby weakening the one-to-one requirement; our sampler
does ensure that each record in \(B\) can be matched to at most one
record in \(A\), but allows for the possiblity that multiple records in
\(B\) match to the same record in \(A\). We explore the ramifications of
this distinction with the El Salvador case study. With this change,
components of the \(\mathbf{Z}\) vector can be computed in parallel,
allowing for significant computational gains. More importantly, since
only the agreement pattern of \(Z_j\) is used for calculations within
the Gibbs sampler, and not the particular record label, we can conduct
this sampling only at the level of the unique agreement patterns,
offering even more computational savings. This will be explained further
in \textcolor{red}{Section XX}.

@@

To address this, we propose a new strategy called \emph{storage
efficient indexing} (SEI). In standard indexing, the modeler decides a
certain criteria that they expect all true matching pairs to satisfy,
and a priori label any record pairs that do not meet that criteria as
non-matching. For example, one might only consider pairs with a certain
similarity score on a field deemed to be important (like first name), or
only pairs with exact matching on a specified number of fields. While
generally chosen to be be quite loose, establishing these criteria
requires knowledge of the problem and invites room for human error. We
propose a method of reducing the comparison space (and reducing storage
requirements) without these drawbacks. Note that all records of the same
agreement pattern have the same probability when sampling \(Z_j\).
Therefore we know that records belonging to an \(h_p\) such that
\(H_{p_j}\) is large are very unlikely to be sampled consistently enough
to be deemed a match through the Bayes estimate, even without
considering the form of the agreement pattern itself. In random
indexing, rather than store all of these unlikely record labels, we
choose to store only a small number \(R\) of them. Posterior
calculations still attribute the appropriate weight to all records
through the summary statistics \(H_{p_j}, H_p^m\), and \(H_p^u\). Rather
than storing \(n_A \times n_B\) record labels, SEI allows us to store at
most \(n_B \times P \times R\) labels, regardless of how large \(n_A\)
is. Coupled with the partitioning strategy for creating comparison
vectors, this greatly expands the size of the linkage tasks we can
undertake.

@@

\hypertarget{simulation-study-1---accuracy}{%
\section{Simulation Study 1 -
Accuracy}\label{simulation-study-1---accuracy}}

In his 2017 paper, Sadinle demonstrated the accuracy of his method by
running \texttt{BRL} on simulated pairs of files with differing levels
of errors and overlap. Each file consists of first name, last name,
gender, postal code, age, and occupation, and following Sadinle's
implementation, we use only four fields for the linkage process: both
names, age, and occupation. Similarity scores for the names are binned
into three levels according the default settings for the
\texttt{compareRecords} function in the \texttt{BRL} package, and age
and occupation similarity scores are binary for exact matching. We use
flat priors for all \(\mathbf{m}\) and \(\mathbf{u}\) parameters, and
run the Gibbs sampler for 1000 iterations, with the first 100 discarded
as burn-in.

Comparing our method against \texttt{BRL} on these same simulated
datasets, we find that our method only has weakened performance in the
most extreme scenario of very high errors and very high overlap across
files. We use the following measures of accuracy in our analysis:
although using only recall and precision is common in the literature, we
find it useful to have a single metric the gauge overall performance,
and thus also include F measure.

@@

\hypertarget{simulation-study-2---speed}{%
\section{Simulation Study 2 - Speed}\label{simulation-study-2---speed}}

To demonstrate speed, we generate comparison vectors from pre-specified
distributions so that we can easily increase the size of the linkage
problem. Distributions are meant to emulate the behavior of similarity
scores across first name, last name, and day, month, and and year of
birth, with 1 indicating an exact match on a particular field and 0
indicating a nonmatch. We simulate these data for different values of
\(n_A\) and \(n_B\), and compare the run-time of \texttt{parlr} against
\texttt{BRL}. Note that the number of unique patterns \(P\) is bounded
above by \(2^5 = 32\), a bound which is consistently attained in the
larger simulations.

We see that at low data size, \texttt{BRL} outperforms, but that
\texttt{parlr} is significantly faster at handling larger data. In
particular, run-time for \texttt{BRL} seems to grow quadratically (or
linearly with the size of both \(A\) and \(B\)) while run-time for
\texttt{parlr} seems to grow linearly (in the size of \(B\)).

@@

The above discussion suggests that for fixed \(n_B\), computation time
should remain mostly constant with growing \(n_A\). Simulation study
suggests that this is true. In the plot below, fixing \(n_B = 500\), we
see linear growth for the run-time under \texttt{BRL} as \(n_A\)
increases, with much more static run-time under \texttt{parlr}. The
slight increases in run-time that we do see are due primarily to the
hashing step, which again can be run in parallel for large data.

@@

the country of El Salvador was immersed in civil war from 1980 to 1991,
and throughout the time, several organizations attempted to document
casualties of the conflict. When estimating the total number of
casualities, one cannot simply sum the numbers recorded by each
organization, as it is likely that the same individuals are recorded in
multiple casuality lists. To obtain a more accurate estimate of the
casualties then, we follow Sadinle's 2017 paper and link records of El
Salvadoran civilian casualties from two sources: El Rescate - Tutela
Regal (ERTL) and the Salvadoran Human Rights Commission (CDHES, by its
acronym in Spanish). The ERTL dataset consists of digitized reports that
had been published throughout the conflict. The CDHES dataset consists
of casualties that had been reported directly to the organization, and
later digitized.

There are several challenges with working with such data. Firstly, both
datasets have been automatically digitized, which inherently leads to
some degree of typographical error. Secondly, the CDHES records are all
second hand accounts reported by individuals, which can result in
additional errors. Lastly, the only fields recorded are given name, last
name, date of death, and place of death; it is relatively common for a
parent and child to share the same given name, resulting in
indistinguishable records for two different individuals. This last point
nearly breaks the above mentioned assumption that there are no
duplicates within files, and reveals a key difference between
\texttt{BRL} and our proposed method.

We only utilized records with nonmissing entries for given and last
name, results in \(n_A = 4420\) files in CHDES and \(n_B = 1323\) files
in ERTL. The names were standardized to account for common misspellings
in the Spanish language, and then compared using a modified Levenstein
distance to account for the fact that second names are often ommited.
Place of birth is recorded by municipality and department within that
municipality; however, since department was missing in 95\% of records
in CHDES and 80\% of records in ERTL, we excluded it from our linkage
process. Thus we conduct linkage using given name, last name,
municipality, and day, month, and year of death. We again use flat
priors for the \(\mathbf{m}\) and \(\mathbf{u}\) parameters, but use
2000 Gibbs iterations with 200 discarded as burn-in to mirror the
implementation in Sadinle 2017.
